Lyman-$\alpha$ photons through rotating outflows

### 2018-08-06 paper/paper.tex
abstract y conclusiones en nueva perspectiva

\begin{abstract}
Outflows and rotation are two ubiquitous kinematic features in the gas
kinematics of galaxies.
Here we introduce a semi-analytic model to quantify how rotating
outflows impact the morphology of the Lyman-$\alpha$ emission line.   
The model is contrasted against Monte Carlo radiative transfer
simulations of outflowing gas with additional solid body rotation.
We explore a range of neutral Hydrogen optical depth of
$10^5\leq\tauh\leq10^7$, rotational velocity $0\leq \vrot/\kms \leq
100$ and outflow velocity $0\leq \vout/\kms\leq 50$.  
We find three important consequences of rotation.
First, it introduces a dependency with viewing angle; second it
produces a line broadening and third it increases the flux at the
line's center.
The semi-analytic model reproduces the radiative transfer results for
the line width and flux change at the line's center within a $7\%$
and $50\%$ precision for an optical depth of
$\tauh=10^5$, respectively, and within $2\%$ and $1\%$ for an optical
depth of $\tauh=10^7$.
The semi-analytic model is a convenient tool to introduce rotational
kinematics as a post-processing step of idealized Monte Carlo
simulations, providing and easy way to include rotation in models that
seek to interpret \lya spectra in systems where rotation is evident
through kinematic maps. 
\end{abstract}

@@

In this paper we explore, for the first time in the literature,
the results of a model for the emergent \lya line from rotating outflows.
The main results for the model are computed from a Monte-Carlo
radiative transfer simulation and confronted with a simple
semi-analytic ansatz that adds the effects of rotation onto results of
pure outflow kinematics.

@@

\begin{itemize}
  \item Inducing a dependency on the viewing angle.
  \item Broadening the line.
  \item Increasing the intensity at the line's center.
\end{itemize}

We find that these effects can be quantitatively explained by a
Doppler shift computed as product of quantities at the surface of last
scattering, namely  $\vec{v}_{\rm rot}\cdot \hatk$, where
$\vec{v}_{\rm rot}$ is the velocity due to rotation and \hatk is
the direction of the photon's propagation.  
These conclusions, specially the confirmation of the Doppler shift  as
a good approximation to model solid body rotation, strengthen the
semi-analytic model reported by \cite{Garavito14} where they modeled
the effect of pure rotation on \lya spectra.

Addressing the second question we posed in the introduction (\emph{to what
extent is it possible to constrain rotational kinematics from the \lya
emission line?})  we can say that as an application to observational data our results suggest that
rotation should also be considered as an explanation of the
non-vanishing \lya intensity at the line's center, which could change
the interpretation of recent observational results
\citep[e.g.][]{2017ApJ...844..171Y,2017A&A...608L...4R} and provide a
test for rotation in the observed systems.

A more straightforward approach takes into account how the
\lya spectra from two different sides of a galaxy can detect
approaching/receding gas motions \citep{Prescott14,2018MNRAS.473.3907A}.
In that case, one has to be careful and take into account that the
peaks of these two spectra is smaller than the expected value of
$2\times\vrot$ due to the different weights at the surface of the
emitting regions.

We also use data from MUSE \citep{2017A&A...606A..12H} to show
to what extent the outputs of the RT results presented here can be
expected to match hundreds of observed \lya spectra, but a full
constraint of rotational properties certainly requires a full Markov
Chain Monte Carlo exploration of parameter space.

Our work confirms that rotation produces features in the \lya line
strong enough that can be expected to be detected in observations and
that should be taken into account into \lya modeling; the Doppler
Shift offers an easy-to-implement approximation to explore such
influence into already existing RT simulations and provide stronger
footing to the interpretation of present and future \lya kinematics
maps \citep[e.g][]{2018MNRAS.473.3907A}.

>>>>

\begin{abstract}
Outflows and rotation are two ubiquitous kinematic features in the gas
kinematics of galaxies. 
Here we introduce a semi-analytic model to quantify how rotating
outflows impact the morphology of the Lyman-$\alpha$ emission line.   
The model is contrasted against Monte Carlo radiative transfer
simulations of outflowing gas with additional solid body rotation.
We explore a range of neutral Hydrogen optical depth of
$10^5\leq\tauh\leq10^7$, rotational velocity $0\leq \vrot/\kms \leq
100$ and outflow velocity $0\leq \vout/\kms\leq 50$.  
We find three important consequences of rotation.
First, it introduces a dependency with viewing angle; second it
produces a line broadening and third it increases the flux at the
line's center.
For all viewing angles, the semi-analytic model reproduces the
radiative transfer results for the line width and flux change at the
line's center within a $7\%$ and $50\%$ precision for an optical depth of
$\tauh=10^5$, respectively, and within $2\%$ and $1\%$ for an optical
depth of $\tauh=10^7$.
The semi-analytic model is a convenient tool to introduce rotational
kinematics as a post-processing step of idealized Monte Carlo
simulations, providing a framework to interpret \lya spectra in
systems where rotation is expected or directly measured through
kinematic maps.  
\end{abstract}

@@

In this paper we explore, for the first time in the literature,
the results of a model for the emergent \lya line from rotating
outflows. 
We use a semi-analytic model, first presented in \citep{Garavito14}, that
to capture the main effects of rotation and confront it against
results from Monte-Carlo radiative transfer simulations. 
The semi-analytic model only takes into account the Doppler shift computed
as product of quantities at the surface of last scattering, namely
$\vec{v}_{\rm rot}\cdot \hatk$, where $\vec{v}_{\rm rot}$ is the
velocity due to rotation and \hatk is the direction of the photon's
propagation.

@@

\begin{itemize}
  \item Inducing a dependency on the viewing angle.
  \item Broadening the line.
  \item Increasing the intensity at the line's center.
\end{itemize}
%
All these effects can be quantitatively explained by the proposed
semi-analytic model.

Addressing the second question we posed in the introduction (\emph{to what
extent is it possible to constrain rotational kinematics from the \lya
emission line?})  we can say that  the most straightforward approach
takes into account how the \lya spectra from two different sides of a
galaxy can detect approaching/receding gas motions
\citep{Prescott14,2018MNRAS.473.3907A}. 
In that case, we also show that one has to be careful and take into
account that the  peaks of these two spectra is smaller than a naively
expected value of $2\times\vrot$ due to the different weights at the
surface of the emitting regions.

As an application to observational data we use data from MUSE
\citep{2017A&A...606A..12H} to show to what extent the outputs of the
RT results presented here can be expected to change the results of
pure outflow simulations and thus bettwe match hundreds of
observed \lya spectra.
We find that the effect on general line morphology statistics
(standard deviation and skewness) is small, although a full constraint
of rotational properties certainly requires a full Markov Chain Monte
Carlo exploration of parameter space.

Our results suggest that rotation should also be considered as an
explanation of the non-vanishing \lya intensity at the line's center,
and become a test for rotation in the observed systems, provided that
the spectral resolution of the instrument is smaller than the expected
rotational velocities.

To summarize, our work confirms that the semi-analytic model based on
the Doppler Shift offers an easy-to-implement approximation to explore
such influence into already existing RT simulations and provide
stronger footing to the interpretation of present and future \lya
kinematics maps \citep[e.g][]{2018MNRAS.473.3907A}.
